---
author: BenS
date: "`r Sys.Date()`"
title: "Theil-Sen Trend Analysis Using Environment Cantebrury Data"
format:
  html:
    self-contained: true
    table-of-contents: true
    toc-location: left
execute:
  warning: false
  message: false
---

```{r}
#| label: setup

knitr::opts_chunk$set(echo = FALSE)
library(tidyverse, warn.conflicts = FALSE)
library(openair)

```

## Introduction

This is an rmarkdown document showing how to calculate 10 year trends calculated using Environment Canterbury data. This is a working document and likely to update and change.

The steps shown are from getting data, cleaning the data (with some decisions on analysis) and finally running the trend calculation. The trends are calculated using the `TheilSen` function from the `openair` package.

::: callout-note

There are some decisions in the code as to what time periods and sites to be included in the trend calculations. These need to be checked and possibly updated each time the trend calculations need to be run.

:::

## Data retrieval

The data retrieval has been put into another function `ecanairquality` which is documented [in a GitHub repository](https://github.com/bws200/ecanairquality).

For this document we can just use the function (which uses the website) to retrieve the data in `.csv` format. The inputs are a start date and end date.

```{r}
#| label: get-data

df_final_long <- ecanairquality::get_daily_all_stations(
  from_date = "01/01/2015",
  to_date = "31/12/2024"
)

```

## Data cleaning and decisions

Now need to get the data in a better state for Theil-Sen analysis. First some reformatting and renaming.

```{r}
#| label: data-cleaning

# Reformat data
df_daily_pm <- df_final_long |>
  # Select only PM10 and PM2.5
  filter(name %in% c("PM10 (ug/m3)", "PM25 (ug/m3)")) |>
  # Reformat to wide
  pivot_wider(
    names_from = name,
    values_from = value
    ) |>
  # "date" column required by openair
  rename(
    date = DateTime,
    PM10 = "PM10 (ug/m3)",
    PM25 = "PM25 (ug/m3)",
    site = StationName
    ) |>
  # Add columns for data checking
  mutate(
    month = lubridate::month(date),
    year = lubridate::year(date),
    days_per_month = lubridate::days_in_month(date)
    ) |> 
  # Add time zone to date
  mutate(date = ymd(date, tz = "Etc/GMT+12"))

```

Before doing the trend estimate it is a good idea to plot the data to check for holes with the function `openair::summaryPlot()`.

```{r}
#| label: pm10-summary-plot-raw
#| column: page-right
#| fig-width: 20
#| fig-height: 30

p_sum_pm10_raw <- select(df_daily_pm, c("date", "site", "PM10")) |>
  openair::summaryPlot(
    pollutant = "PM10",
    type = "site",
    fontsize = 20
  )

```

And PM~2.5~

```{r}
#| label: pm25-summary-plot-raw
#| column: page-right
#| fig-width: 20
#| fig-height: 30

p_sum_pm25_raw <- select(df_daily_pm, c("date", "site", "PM25")) |> 
  openair::summaryPlot(
    pollutant = "PM25",
    type = "site",
    fontsize = 20
  )


```
The  summary plots show some decisions need to be made as to which sites to include and which periods should be used. The selection of periods are based on the [trend information sheet](https://boplass2.sharepoint.com/:w:/r/sites/NationalAirQualityGrp/_layouts/15/Doc.aspx?sourcedoc=%7BCD47641A-F0E7-420C-B974-40EF61C84B4D%7D&file=LAWA%20air%20quality%20trends%20guidance%2013022022.docx&_DSL=1&action=default&mobileredirect=true). The main decision point is that recent data should be discarded and trends calculated from the period prior if there are large holes (> 5 months) in the data.

The following is a list of the decisions made. The code should reflect these decisions as well.

- No trend should be calculated for Washdyke as the site has been changed and there has been a significant change in concentrations and occurrence of exceedances so we can remove those Washdyke stations from the data set.
- For St Albans, the site was shut down on 2020-11-11 so that will be the end date for this trend analysis for St Albans. There are still six consecutive years (just!) at the old Coles Place site which is enough for trend analysis.

```{r}
#| label: data-additional-decisions

# These need updating each year trends are calculated

df_daily_pm_clean <- df_daily_pm |>
  # Remove Washdyke Flat Road and Waimate Stadium site
  filter(!(site %in% c("Washdyke Flat Road", "Waimate Stadium"))) |> 
  
  # Remove St Albans data prior to this date due to site movement. There are still six consecutive years with over 75%
  filter(site != "St Albans EP")

```

```{r}
#| label: pm10-summary-plot-clean
#| column: page-right
#| fig-width: 20
#| fig-height: 30

p_sum_pm10_clean <- select(df_daily_pm_clean, c("date", "site", "PM10")) |> 
  openair::summaryPlot(
    pollutant = "PM10",
    type = "site",
    fontsize = 20
  )

```

And PM~2.5~

```{r}
#| label: pm25-summary-plot-clean
#| column: page-right
#| fig-width: 20
#| fig-height: 30

p_sum_pm25_clean <- select(df_daily_pm_clean, c("date", "site", "PM25")) |>
  openair::summaryPlot(
    pollutant = "PM25",
    type = "site",
    fontsize = 20
  )


```




## Theil-Sen analysis

So now we can run the Theil-Sen trend analysis on the PM~10~ and PM~2.5~ datasets

```{r}
#| label: trend-analysis-pm10
#| fig-width: 20
#| fig-height: 20
#| column: page-right

# PM10 trend analysis

p_pm10 <- openair::TheilSen(
  df_daily_pm_clean,
  ylab = "PM10 (ug/m3)",
  pollutant = "PM10",
  deseason = TRUE,
  date.format = "%Y",
  type = "site",
  data.thresh = 75,
  fontsize = 30
  )

```

```{r}
#| label: trend-analysis-pm25
#| fig-width: 20
#| fig-height: 20
#| column: page-right

# And TheilSen for PM2.5
p_pm25 <- openair::TheilSen(
  df_daily_pm_clean,
  ylab = "PM2.5 (ug/m3)",
  pollutant = "PM25",
  deseason = TRUE,
  date.format = "%Y",
  type = "site",
  data.thresh = 75,
  fontsize = 30
  )


```

## Data export

Here the data is exported to show what has been used in the trend analysis calculations for comparison or to reproduce somewhere else without running the data grab from the website.

```{r}
#| label: data-export

df_final_wide <- pivot_wider(df_final_long, names_from = "name", values_from = "value")

write.csv(
  df_final_wide,
  file = paste0("data_export_",Sys.Date() ,".csv")
  )

# Note most of the data worked with is in the long format
head(df_final_long)
# as oppposed to the wide format which is exported here
head(df_final_wide)

```


## Trend analysis using `.csv` file

We can also run the same analysis using data retrieved from a .csv file. This is handy for comparison.

::: callout-note
Note the dataset dates need to match if comparing .csv read and retrievals from the website
:::

### Get the data

This assumes the data has been prepared in a .csv file which is included in this [GitHub](https://github.com/bws200/theil-sen-trend-analysis-ecan) repository under `/data`. It is for Environment Canterbury's Rangiora site only. A larger .csv file could be created which includes site name as an additional column.

```{r}
#| label: data-read-csv
#| column: page-right

# Read file
df_csv <- read.csv(
  file = "data/Rangiora PM10 2013 to 2022.csv"
) |>
  rename(
    date = DateTime,
    PM10 = PM10..ug.m3.,
    PM25 = PM2.5..ug.m3.
  ) |>
  mutate(date = lubridate::dmy(date)) |>
  filter(date >= "2016-01-01")

df_csv |> 
  # Add in time zone to date
  mutate(date = ymd(date, tz = "Etc/GMT+12")) |> 
  openair::summaryPlot(
    pollutant = "PM25"
  )

```

::: callout-note
The `.csv` dataset is not complete for 2022
:::

```{r}
#| label: trend-analysis-pm10-csv
#| column: page-right

# Run the trend analysis
openair::TheilSen(
  df_csv,
  pollutant = "PM25",
  deseason = TRUE,
  date.format = "%Y",
  data.thresh = 75
)

```

## Appendix

### St Albans trend check

As the St Albans set of data uses a trend calculation with 6 years of data prior to moving to the new site, we will run some trend calculation checks on other scenarios.

```{r}
#| label: st-albans
#| column: page-right
#| fig-width: 20
#| fig-height: 10

# df_daily_pm_st_albans <-

df_stalbans <- df_daily_pm |>
  filter(site == "St Albans")

df_stalbansEP <- df_daily_pm |> 
  filter(site == "St Albans EP")

df_stalbans_combined <- df_daily_pm |> 
  filter(site %in% c("St Albans", "St Albans EP"))

df_stalbans_combined$site <- "St Albans Combined"

df_stalbans_combined_limited <- df_stalbans_combined

df_stalbans_combined_limited$site <- "St Albans Combined Limited"

df_stalbans_combined_2017 <- df_stalbans_combined_limited |> 
  filter(between(year, 2017, 2023)) |> 
  mutate(site = "St Albans 2017")

df_stalbans_combined_2018 <- df_stalbans_combined_limited |> 
  filter(between(year, 2018, 2024)) |> 
  mutate(site = "St Albans 2018")

df_stalbans_combined_2019 <- df_stalbans_combined_limited |> 
  filter(between(year, 2019, 2025)) |> 
  mutate(site = "St Albans 2019")

df_daily_pm_st_albans_all <- rbind(df_stalbans, df_stalbansEP, df_stalbans_combined, df_stalbans_combined_2017, df_stalbans_combined_2018, df_stalbans_combined_2019)

df_daily_pm_st_albans_all |> 
  openair::summaryPlot(
    pollutant = "PM10",
    type = "site",
    fontsize = 20
  )

df_daily_pm_st_albans_all |> 
  select(date, site, PM10) |>
  openair::TheilSen(
    ylab = "PM10 (ug/m3)",
    pollutant = "PM10",
    deseason = TRUE,
    date.format = "%Y",
    type = "site",
    data.thresh = 75,
    fontsize = 30
  )
```



### Rounding

The standard round function in R doesn't work like others in excel so another function needs to be added. See this [Stack Overflow discussion](https://stackoverflow.com/questions/12688717/round-up-from-5) for more discussion on rounding in R.

```{r}
#| label: data-note-on-rounding
#| include: true

# Round 0.5 is expected to round to 1 but R rounds it down to 0
round(0.5)

# But instead we need a function to round like excel e.g. Function to always round 0.5 up
round2 <- function(x, digits = 0) {
  posneg <- sign(x)
  z <- abs(x) * 10^digits
  z <- z + 0.5
  z <- trunc(z)
  z <- z / 10^digits
  z * posneg
}

# And try again with round2 which should round to 1
round2(0.5)

# An easier method is just using the in-built janitor::round_half_up function which implements the same round2 function but without defining it
janitor::round_half_up(0.5)

```
### `openair` requirements

Noting here as tz information was found to be required for `openair::summaryPlot()` to display correctly. E.g.

```{r}
#| eval: false

mutate(date = ymd(date, tz = "Etc/GMT+12"))
```




